Long term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-20 article/article.Rmd
Commit 2024-06-20 17:13
Commit 2024-06-20 20:14

<!-- Radiation data quality check -->
To further improve the quality of the irradiance data, a manual screening was
performed, in order to remove inconsistent and erroneous recordings that can occur
stochastically or systematically during the long operation of the instrument.  The
manual screening was aided by a radiation data quality assurance procedure, adjusted
for the site, which was based on the methods of Long and Shi\ [@Long2006;
@Long2008a].  Thus, problematic recordings have been excluded from further
processing.  Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality of the
radiation measurements used in this study.
<!-- Data selection -->
To preserve an unbiased representation of the data we applied a constraint, similar
the one used by @CastillejoCuberos2020. For each valid hour of day, there must exist
at least 45 minutes of valid measurements, including nighttime, when near sunrise and
sunset. Days with less than 5 valid hours were rejected completely.
<!--# CastillejoCuberos2020 There must be at least 5 h of valid data (Tests 1–7),
otherwise all data for this day is flagged as invalid. There must be at least This
work valid data 45 min of valid data (Tests 1–7) for an hour to be flagged as valid,
otherwise all minute data for this hour is flagged as invalid. -->
Furthermore, due to the significant measurement uncertainty in GHI when the Sun is
near the horizon, and due to some systematic obstructions by nearby buildings, we
have excluded all measurements with solar zenith angle (SZA) greater than $`r 90 -
BIO_ELEVA`^\circ$.

@@

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute $\text{GHI}_i$ from the $\text{GHI}_\text{ref}$ corresponding to
cloud-free atmosphere. 
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023], and
the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.  We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the CE events. However, by inspecting the daily plots of
irradiance it became evident that changes in atmospheric conditions introduced
numerous false positive and false negative results.  The main reason for these
discrepancies is the variability of the effects of aerosols and water vapor which
were not taken into account in the two simple methods.  To produce a more
representative reference  we included the effects of these factors through variables
using a radiative transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}..

@@

<!-- Aerosol change -->
The above clear sky reference values are based on the climatological AOD and WC,
hence they cannot describe accurately the long term variation of
$\text{GHI}_\text{ref}$ due to changes in the two atmospheric constituents, mainly
AOD.  As reported by @Natsis2023, there is a long term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky data was
attributed to long-term changes in aerosol effects.  Therefore, an adjustment of the
$\text{GHI}_\text{ref}$ used in this study is necessary. As AERONET data start only
in 2003, we used for the period 1993 -- 2005 estimates of changes in AOD at
$340\,\text{nm}$ derived from a collocated Brewer spectrophotometer [@Kazadzis2007]
to calculate the trend in $\text{GHI}_\text{ref}$ during this period.

@@

To create a unified trend for the long term change of the clear sky irradiance, we
simulated the values of AOD derived from those sources with Libradtran. For both
inputs, we used the AOD at $500\,\text{nm}$, which was inferred by the available
Ångström coefficients. We choose the SZA of $55^\circ$ as a representative value for
all the runs.

According to this study, in the period 1997 -- 2005 the mean AOD at $340\,\text{nm}$
$0.403$ with a trend of $-3.8\pm0.93\,\%$ per year, corresponding to a change of
$0.0153$ per year.  Using an Ångström coefficient $\alpha = 1.6$, this translates to
a change in the Ångström coefficient $\beta=0.00272$ per year (or $\beta=0.084$ in
1997 and $\beta=0.059$ in 2005).  Simulations with uvspec for the above Ångström
coefficients, with WC of
$`r round(BR[year==1997,pw_avg_mm],1)`\,\text{mm}$ and 
$`r round(BR[year==2005,pw_avg_mm],1)`\,\text{mm}$
for 1997 and 2005 respectively and for a SZA of $55^\circ$ reveal a trend of
$+0.21\,\%$ per year in $\text{GHI}_\text{ref}$. 
The SZA of $55^\circ$ was chosen as representative of all days in the year in order
to get a rough estimate of the annually averaged change in clear sky irradiance.  For
the period 2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET in
a similar simulation scheme to calculate the monthly mean of clear sky irradiance,
and finally the trend of $+0.14\,\%$ per year. We applied these two long term changes
(see Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
$\text{GHI}_\text{ref}$, in order to create a more realistic representation of the
clear-sky irradiance for the whole period of study.

@@

First, we evaluated the performance of the modelled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.
<!-- $\tau_{\text{cs}} = \tau_{500\text{nm}} - 1\sigma$ and $w_{h\text{cs}} = w_h - 1\sigma$
($\text{GHI}_\text{CSm}$). -->
These values represent typical atmosphere in Thessaloniki with lower than average
load of aerosols and humidity, which are the main factor that attenuate the GHI,
excluding clouds.  With this approach the simulated $\text{GHI}_\text{ref}$ should be
generally greater than the measured GHI when aerosols are more abundant.  To
compensate for this, we increased the $\text{GHI}_\text{ref}$ by 
($`r 100*(C4_cs_ref_ratio-1)`\,\%$)
with an additional constant offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$,
as described in Equation\nobreakspace\ref{eq:CE4}.
\begin{equation}
\text{CE} : \text{E}_\text{i} > `r C4_GLB_diff_THRES` + `r C4_cs_ref_ratio` \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}
where: $\text{E}$ the measured irradiance, $\text{E}_\text{CSm}$ the selected
modelled clear sky irradiance, and $i$ each of the one-minute observation.  This is
the criterion of our CE identification.

These values were determined through the implementation of an empirical method with
manual inspection of the CE identification results  on selected days of the whole
dataset.  We used seven sets of days with characteristics relevant to the efficiency
of the identification threshold.  These sets were random groups of about 20 -- 30
days with the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of EC events,
(e) with at least $60\,\%$ of the day length without clouds and presence of EC events,
(h) randomly selected days, and
(i) manually selected days.
Where needed in some of the edge cases, we also used images from a sky-camera
to further aid the decision of the manual inspection.

@@

<!-- extreme cases criterion -->
A sub-category  of the CE events that is often discussed in the relevant bibliography
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for the actual Sun -- Earth distance, and $i$ each of the one-minute observation.

>>>>

<!-- Radiation data quality check -->
To further improve the quality of the irradiance data, a manual screening was
performed, in order to remove inconsistent and erroneous recordings that can occur
stochastically or systematically during the long operation of the instrument.  The
manual screening was aided by a radiation data quality assurance procedure, adjusted
for the site, which was based on the methods of Long and Shi\ [@Long2006;
@Long2008a].  Thus, problematic recordings have been excluded from further
processing.
<!--
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality of the
radiation measurements used in this study.
-->
<!-- Data selection -->
<!--
To preserve an unbiased representation of the data we applied a constraint, similar
the one used by @CastillejoCuberos2020. For each valid hour of day, there must exist
at least 45 minutes of valid measurements, including nighttime, when near sunrise and
sunset. Days with less than 5 valid hours were rejected completely.
-->
<!--
# CastillejoCuberos2020 There must be at least 5 h of valid data (Tests 1–7),
otherwise all data for this day is flagged as invalid. There must be at least This
work valid data 45 min of valid data (Tests 1–7) for an hour to be flagged as valid,
otherwise all minute data for this hour is flagged as invalid.
-->
Furthermore, due to the significant measurement uncertainty in GHI when the Sun is
near the horizon, and due to some systematic obstructions by nearby buildings, we
have excluded all measurements with solar zenith angle (SZA) greater than $`r 90 -
BIO_ELEVA`^\circ$.

@@

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute GHI from the $\text{GHI}_\text{ref}$ corresponding to
cloud-free atmosphere.
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023], and
the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.  We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the CE events. However, by inspecting the daily plots of
irradiance it became evident that changes in atmospheric conditions introduced
numerous false positive and false negative results.  The main reason for these
discrepancies is the variability of the effects of aerosols and water vapor which
were not taken into account in the two simple methods.  To produce a more
representative reference  we included the effects of these factors through variables
using a radiative transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}..

@@

<!-- Aerosol change -->
The above clear sky reference values are based on the climatological AOD and WC,
hence they cannot describe accurately the long term variation of
$\text{GHI}_\text{ref}$ due to changes in the two atmospheric constituents, mainly
AOD.  As reported by @Natsis2023, there is a long term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky data was
attributed to long-term changes in aerosol effects.  Therefore, an adjustment of the
$\text{GHI}_\text{ref}$ during the period of study was made using simulation with the
RTM based on the long-term variations of the AOD. As AERONET data start only in 2003,
we used for the period 1993 -- 2005 estimates of changes in AOD at $340\,\text{nm}$
derived from a collocated Brewer spectrophotometer [@Kazadzis2007] to calculate the
trend in $\text{GHI}_\text{ref}$ during this period.

@@

<!--
To create a unified trend for the long term change of the clear sky irradiance, we
simulated the values of AOD derived from those sources with Libradtran. For both
inputs, we used the AOD at $500\,\text{nm}$, which was inferred by the available
Ångström coefficients. We choose the SZA of $55^\circ$ as a representative value for
all the runs.
-->

According to this study, in the period 1997 -- 2005 the mean AOD at $340\,\text{nm}$
is $0.403$ with a trend of $-3.8\pm0.93\,\%$ per year, corresponding to a change of
$0.0153$ per year.  Using an Ångström coefficient $\alpha = 1.6$, this translates to
a change in the Ångström coefficient $\beta=0.00272$ per year (or $\beta=0.084$ in
1997 and $\beta=0.059$ in 2005).  Simulations with uvspec for the above Ångström
coefficients, with WC of
$`r round(BR[year==1997,pw_avg_mm],1)`\,\text{mm}$ and 
$`r round(BR[year==2005,pw_avg_mm],1)`\,\text{mm}$
for 1997 and 2005 respectively and for a SZA of $55^\circ$ reveal a trend of
$+0.21\,\%$ per year in $\text{GHI}_\text{ref}$. 
The SZA of $55^\circ$ was chosen as representative of all days in the year in order
to get a rough estimate of the annually averaged change in clear sky irradiance.  For
the period 2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET in
a similar simulation scheme to calculate the monthly mean of clear sky irradiance,
and finally the trend of $+0.14\,\%$ per year. We applied these two long term changes
(see Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
$\text{GHI}_\text{ref}$, in order to create a more realistic representation of the
clear-sky irradiance for the whole period of study.

@@

First, we evaluated the performance of the modelled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.
<!-- $\tau_{\text{cs}} = \tau_{500\text{nm}} - 1\sigma$ and $w_{h\text{cs}} = w_h - 1\sigma$
($\text{GHI}_\text{CSm}$). -->
These values represent typical atmospheres in Thessaloniki with lower than average
load of aerosols and humidity, which are the main factors that attenuate the GHI,
excluding clouds.  With this approach the simulated $\text{GHI}_\text{ref}$ should be
generally greater than the measured GHI when aerosols are more abundant.  To
compensate for this, we increased the $\text{GHI}_\text{ref}$ by 
($`r 100*(C4_cs_ref_ratio-1)`\,\%$)
with an additional constant offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$,
as described in Equation\nobreakspace\ref{eq:CE4}.
\begin{equation}
\text{CE} : \text{E}_\text{i} > `r C4_GLB_diff_THRES` + `r C4_cs_ref_ratio` \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}
where: $\text{E}$ the measured irradiance, $\text{E}_\text{CSm}$ the selected
modelled clear sky irradiance, and $i$ each of the one-minute observation.  This is
the criterion of our CE identification.

These values were determined through the implementation of an empirical method with
manual inspection of the CE identification results  on selected days of the whole
dataset.  We used seven sets of days with characteristics relevant to the efficiency
of the identification threshold.  These sets were random groups of about 20 -- 30
days with the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of EC events,
(e) with at least $60\,\%$ of the day length without clouds and presence of EC events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed in some of the edge cases, we also used images
from a sky-camera to further aid the decision of the manual inspection.

@@

<!-- extreme cases criterion -->
A sub-category  of the CE events that is often discussed in the relevant literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere. In this case the threshold EEEEEEEEEE:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for the actual Sun -- Earth distance, and $i$ each of the one-minute observation.

### 2024-06-20 article/article.tex
Commit 2024-06-20 17:13
Commit 2024-06-20 20:14

To further improve the quality of the irradiance data, a manual
screening was performed, in order to remove inconsistent and erroneous
recordings that can occur stochastically or systematically during the
long operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality
of the radiation measurements used in this study. To preserve an
unbiased representation of the data we applied a constraint, similar the
one used by \citet{CastillejoCuberos2020}. For each valid hour of day,
there must exist at least 45 minutes of valid measurements, including
nighttime, when near sunrise and sunset. Days with less than 5 valid
hours were rejected completely. Furthermore, due to the significant
measurement uncertainty in GHI when the Sun is near the horizon, and due
to some systematic obstructions by nearby buildings, we have excluded
all measurements with solar zenith angle (SZA) greater than
\(78^\circ\).

@@

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute \(\text{GHI}_i\) from
the \(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere.
First, we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}, and the total solar irradiance (TSI) at
the top of the atmosphere, adjusted for the Sun-Earth distance. We have
tested both methods by using an appropriate relative threshold and/or an
additional constant offset. The initial results showed that we can
detect a big portion of the CE events. However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric
conditions introduced numerous false positive and false negative
results. The main reason for these discrepancies is the variability of
the effects of aerosols and water vapor which were not taken into
account in the two simple methods. To produce a more representative
reference we included the effects of these factors through variables
using a radiative transfer model (RTM). The applied methodology is
discussed in section\nobreakspace{}\ref{rtmcs}..

@@

The above clear sky reference values are based on the climatological AOD
and WC, hence they cannot describe accurately the long term variation of
\(\text{GHI}_\text{ref}\) due to changes in the two atmospheric
constituents, mainly AOD. As reported by \citet{Natsis2023}, there is a
long term brightening effect in the GHI data of Thessaloniki for the
period 1993 -- 2023, which for clear-sky data was attributed to
long-term changes in aerosol effects. Therefore, an adjustment of the
\(\text{GHI}_\text{ref}\) used in this study is necessary. As AERONET
data start only in 2003, we used for the period 1993 -- 2005 estimates
of changes in AOD at \(340\,\text{nm}\) derived from a collocated Brewer
spectrophotometer \citep{Kazadzis2007} to calculate the trend in
\(\text{GHI}_\text{ref}\) during this period.

To create a unified trend for the long term change of the clear sky
irradiance, we simulated the values of AOD derived from those sources
with Libradtran. For both inputs, we used the AOD at \(500\,\text{nm}\),
which was inferred by the available Ångström coefficients. We choose the
SZA of \(55^\circ\) as a representative value for all the runs.

According to this study, in the period 1997 -- 2005 the mean AOD at
\(340\,\text{nm}\) \(0.403\) with a trend of \(-3.8\pm0.93\,\%\) per
year, corresponding to a change of \(0.0153\) per year. Using an
Ångström coefficient \(\alpha = 1.6\), this translates to a change in
the Ångström coefficient \(\beta=0.00272\) per year (or \(\beta=0.084\)
in 1997 and \(\beta=0.059\) in 2005). Simulations with uvspec for the
above Ångström coefficients, with WC of \(15.7\,\text{mm}\) and
\(15.4\,\text{mm}\) for 1997 and 2005 respectively and for a SZA of
\(55^\circ\) reveal a trend of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean of clear
sky irradiance, and finally the trend of \(+0.14\,\%\) per year. We
applied these two long term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), in order to create a more realistic
representation of the clear-sky irradiance for the whole period of
study.

@@

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmosphere in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factor
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
increased the \(\text{GHI}_\text{ref}\) by (\(4\,\%\)) with an
additional constant offset of \(15\,\text{W}/\text{m}^2\), as described
in Equation\nobreakspace\ref{eq:CE4}. \begin{equation}
\text{CE} : \text{E}_\text{i} > 15 + 1.04 \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} where: \(\text{E}\) the measured irradiance,
\(\text{E}_\text{CSm}\) the selected modelled clear sky irradiance, and
\(i\) each of the one-minute observation. This is the criterion of our
CE identification.

These values were determined through the implementation of an empirical
method with manual inspection of the CE identification results on
selected days of the whole dataset. We used seven sets of days with
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. Where needed in some of the edge cases, we also used images from a
sky-camera to further aid the decision of the manual inspection.

@@

A sub-category of the CE events that is often discussed in the relevant
bibliography \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance exceeds the TSI at the top of the
atmosphere: \begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation} where: \(\theta\) the solar zenith angle, \(E_{\odot}\)
the solar constant adjusted for the actual Sun -- Earth distance, and
\(i\) each of the one-minute observation.

>>>>

To further improve the quality of the irradiance data, a manual
screening was performed, in order to remove inconsistent and erroneous
recordings that can occur stochastically or systematically during the
long operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\).

@@

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}, and the total solar irradiance (TSI) at
the top of the atmosphere, adjusted for the Sun-Earth distance. We have
tested both methods by using an appropriate relative threshold and/or an
additional constant offset. The initial results showed that we can
detect a big portion of the CE events. However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric
conditions introduced numerous false positive and false negative
results. The main reason for these discrepancies is the variability of
the effects of aerosols and water vapor which were not taken into
account in the two simple methods. To produce a more representative
reference we included the effects of these factors through variables
using a radiative transfer model (RTM). The applied methodology is
discussed in section\nobreakspace{}\ref{rtmcs}..

@@

The above clear sky reference values are based on the climatological AOD
and WC, hence they cannot describe accurately the long term variation of
\(\text{GHI}_\text{ref}\) due to changes in the two atmospheric
constituents, mainly AOD. As reported by \citet{Natsis2023}, there is a
long term brightening effect in the GHI data of Thessaloniki for the
period 1993 -- 2023, which for clear-sky data was attributed to
long-term changes in aerosol effects. Therefore, an adjustment of the
\(\text{GHI}_\text{ref}\) during the period of study was made using
simulation with the RTM based on the long-term variations of the AOD. As
AERONET data start only in 2003, we used for the period 1993 -- 2005
estimates of changes in AOD at \(340\,\text{nm}\) derived from a
collocated Brewer spectrophotometer \citep{Kazadzis2007} to calculate
the trend in \(\text{GHI}_\text{ref}\) during this period.

According to this study, in the period 1997 -- 2005 the mean AOD at
\(340\,\text{nm}\) is \(0.403\) with a trend of \(-3.8\pm0.93\,\%\) per
year, corresponding to a change of \(0.0153\) per year. Using an
Ångström coefficient \(\alpha = 1.6\), this translates to a change in
the Ångström coefficient \(\beta=0.00272\) per year (or \(\beta=0.084\)
in 1997 and \(\beta=0.059\) in 2005). Simulations with uvspec for the
above Ångström coefficients, with WC of \(15.7\,\text{mm}\) and
\(15.4\,\text{mm}\) for 1997 and 2005 respectively and for a SZA of
\(55^\circ\) reveal a trend of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean of clear
sky irradiance, and finally the trend of \(+0.14\,\%\) per year. We
applied these two long term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), in order to create a more realistic
representation of the clear-sky irradiance for the whole period of
study.

@@

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factors
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
increased the \(\text{GHI}_\text{ref}\) by (\(4\,\%\)) with an
additional constant offset of \(15\,\text{W}/\text{m}^2\), as described
in Equation\nobreakspace\ref{eq:CE4}. \begin{equation}
\text{CE} : \text{E}_\text{i} > 15 + 1.04 \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} where: \(\text{E}\) the measured irradiance,
\(\text{E}_\text{CSm}\) the selected modelled clear sky irradiance, and
\(i\) each of the one-minute observation. This is the criterion of our
CE identification.

These values were determined through the implementation of an empirical
method with manual inspection of the CE identification results on
selected days of the whole dataset. We used seven sets of days with
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed in some of the edge cases, we
also used images from a sky-camera to further aid the decision of the
manual inspection.

@@

A sub-category of the CE events that is often discussed in the relevant
literature \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance exceeds the TSI at the top of the
atmosphere. In this case the threshold EEEEEEEEEE: \begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation} where: \(\theta\) the solar zenith angle, \(E_{\odot}\)
the solar constant adjusted for the actual Sun -- Earth distance, and
\(i\) each of the one-minute observation.
